---
title: "Reproducible Research: Peer Assessment 1"
keep_md: yes
output: md_document
date: "February 7, 2016"
---

<B>Assignment Part 1:</B> Loading and preprocessing the data for project 1  

```{r}
curPackages <- installed.packages()[,1]
reqPackages <- c("data.table","dplyr","lubridate","sqldf")
for (pkg in reqPackages){
     if (!is.element(pkg,curPackages))
          install.packages(pkg)
}

# Load the required packages if necessary
library(data.table)
library(dplyr)
library(ggplot2)
library(lubridate)

# Load the dataset
steps_raw <- fread("activity.csv",header=TRUE)
steps_dt <- na.omit(steps_raw) %>% # Remove all rows with NA in steps column
     mutate(date=ymd(date))        # Change the character date to POSIX date

```

<B>Assignment Part 2:</B> Calculate the mean total number of steps taken per day.  
<B>Task 1:</B> Calculate the total number of steps taken per day  

```{r}
steps_total <- group_by(steps_dt,date) %>% summarise(total_steps = sum(steps))
summary(steps_total)
```

<B>Task 2:</B> Make a histogram of the total number of steps taken each day  

```{r}
hist(steps_total$total_steps,col="green", 
     xlab="Total Steps",
     main="Total Steps Frequency Occurrence\n(Across all days)")
     
```

<B>Task 3:</B> Calculate and report the mean and median of the total number of steps taken per day  

```{r}
mean_steps <- mean(steps_total$total_steps)
mean_steps
median_steps <- median(steps_total$total_steps)
median_steps
```
For visual clarity, the following code displays a boxplot of the steps showing the median steps along with the calculated mean steps overlaid with a red point.

```{r}
main="Box plot for totals steps with both Mean and Median shown"
xLabel1 <- paste("Total Steps with Median",median_steps,sep='=')
xLabel2 <- paste("\nMean steps calculated across all days displayed in red",round(mean_steps),sep='=')
allLabel<- paste(xLabel1,xLabel2)

boxplot(steps_total$total_steps, col = "blue", xlab=allLabel,main=main)
points(c(mean_steps),col="red",lwd=6)
```

<B>Assignment Part 3:</B> Calculate the average daily activity patterns.  

<B>Task 1:</B> Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r}
steps_ave <- group_by(steps_dt,interval) %>% 
     summarise(ave_steps = round(mean(steps)))    
summary(steps_ave)

qplot(interval,
      ave_steps,
      data=steps_ave,
      geom="line",
      xlab="5 min Intervals (represented as 0 - 2400 hrs)",
      ylab="Average Steps per 5 min Interval",
      main="Average Number of Steps\n(calculated across all days)")
```

<B>Task/Question 2:</B> Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
<B>Answer:</B> According to the data below interval 835 contains the maximum average number steps of 206.  

```{r}
interval_dt = filter(steps_ave,ave_steps=which.max(steps_ave$ave_steps))
interval_dt
print(paste("Interval which contains the maximum number of steps: ",interval_dt$interval,sep=" "))
print(paste("The maxium average at that interval: ",interval_dt$ave_steps,sep=" "))
```

For additional visual clarity, the following code displays the average steps across all days at each interval and highlights the interval at which the average max steps are taken:  

```{r}
xLabel1 <- "5 min Intervals (represented as 0 - 2400 hrs)"
xLabel2 <- paste("\nMax steps occur at interval",interval_dt$interval,sep='=')
allLabel <- paste(xLabel1,xLabel2)
qplot(interval,
      ave_steps,
      data=steps_ave,
      geom="line",
      xlab=allLabel,
      ylab="Average Steps per 5 min Interval\n",
      main="Average Number of Steps (calculated across all days)") +
      geom_vline(xintercept=interval_dt$interval, 
                 linetype="dashed",  color = "red",lwd=2)
```

<B>Assignment Part 4:</B> Imputing missing values  

The following code replaces all of the missing (i.e. NA) values with the average value for the correspinding intervals previously calculated.  

<B>Task 1:</B> Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  

```{r}
num_complete <- nrow(steps_dt)
missing_values <- nrow(steps_raw) - num_complete
missing_values
```

<B>Task 2:</B> Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.   

and...

<B>Task 3:</B> Create a new dataset that is equal to the original dataset but with the missing data filled in.   

```{r}
steps_filled_dt <- mutate(steps_raw,
                          steps=ifelse(is.na(steps),
                          steps_ave$ave_steps,
                          steps_raw$steps)) %>% mutate(date=ymd(date))

summary(steps_filled_dt)
```

<B>Task 4:</B> Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.  

<B>Question:</B> Do these values differ from the estimates from the first part of the assignment?  
<B>Answer:</B> Yes they do differ but only slightly and seem to have negligible impact.  

```{r}
steps_total_filled <- group_by(steps_filled_dt,date) %>% summarise(total_steps = sum(steps))       
hist(steps_total_filled$total_steps,col="green", 
     xlab="Total Steps",
     main="Total Steps Frequency Occurrence with NA's Replaced\n(Across all days)")
```

<B>Question:</B> What is the impact of imputing missing data on the estimates of the total daily number of steps?    
<B>Answer:</B>From the data below it appears that the imputation of the average for each interval to replace all NA values has verily little impact on the median and mean steps taken across all days.  

```{r}
mean_steps_filled <- mean(steps_total_filled$total_steps)
median_steps_filled <- median(steps_total_filled$total_steps)

print(paste("Mean/Median steps calculated with NA omitted:  ",paste(round(mean_steps),median_steps,sep="/"),sep=" "))
print(paste("Mean/Median steps calculated with NA Filled-in: ",paste(round(mean_steps_filled),median_steps_filled,sep="/"),sep=" "))
```

The following code creates comparative boxplots for omitted and filled in NA values.  
As you can see from this chart, the impact of replacing the NA values with the average daily values calculated at each interval appears to be small.  

```{r}
main="Box plots for totals steps for both NA's omitted and filled-in\n(with both Mean and Median shown)"
xLabel1 <-"Total Steps with Median"
xLabel2 <- "\nMean steps calculated across all days for each NA group displayed in red"
allLabel<- paste(xLabel1,xLabel2)

boxplot(steps_total$total_steps,
        steps_total_filled$total_steps,
        names=c("NA Omitted","NA Filled-in"),
        col=c("yellow","blue"),
        xlab=allLabel,
        main=main)
points(c(mean_steps,mean_steps_filled),col="red",lwd=6)


```

<B>Assignment Part 5:</B> Weekday/Weekend Comparative Analysis  
<B>Question:</B> Are there differences in activity patterns between weekdays and weekends?  
<B>Answer:</B> Yes, there appears to be more activity (i.e. greater number of steps) during the weekdays in the morning hours but on average more steps are taken on the weekends later in the day.

<B>Task 1:</B> Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  

```{r}
steps_wDaytype <- mutate(steps_filled_dt, 
                       dayType=factor(ifelse(weekdays(date) %in% c("Saturday","Sunday"),
                                    "weekend",
                                    "weekday")))
str(steps_wDaytype)
```

<B>Task 2:</B> Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  

```{r}
steps_ave_b <- group_by(steps_wDaytype,interval,dayType) %>%    # Group dataset by 5 min interval
     summarise(ave_steps = round(mean(steps)))                  # Calculate the average across all intervals

qplot(interval,
      ave_steps,
      data=steps_ave_b,
      facets= dayType ~ .,
      geom="line",
      xlab="5 min Intervals (represented as 0 - 2400 hrs)",
      ylab="Average Steps per 5 min Interval",
      main="Average Number of Steps Comparison\nWeekdays vs. Weekends")

```
